decompr: GVC decomposition in R

### 2015-02-10 draft.Rnw
stelina's comments

\section{Introduction}
Global Value Chains (GVCs) refer to the quickly expanding internationalization of production networks.
Most goods we use nowadays consist of parts that are sourced from different corners of the planet 
and are assembled across different continents. A popular example of this development is the Iphone,
which uses inputs from at least five countries (USA, China, Germany, Taiwan, South Korea) and is assembled in two (USA and China).
This has made GVCs a central topic in research on trade and development policy.
Both policy makers and academia increasingly value the growth opportunities GVCs offer to global trade and, 
especially, to developing countries.
However, analysing this phenomenon empirically requires complex matrix manipulations, 
since the relevant data is only available in the form of gross flows.
The decompr package enables researchers with little background in matrix algebra and linear programming to easily derive standard GVC indicators for statistical analysis.

@@

\begin{itemize}
 \item \verb!load_tables_vectors!; transforms the input objects to an object used for the decompositions (class: decompr)
 \item \verb!leontief!; takes a decompr object and applies the Leontief decomposition
 \item \verb!wwz!; takes a decompr object and applies the Wang-Wei-Zhu decomposition
 \item \verb!vertical_specialisation!; takes a decompr object and returns a measure of vertical specialisation (also works with \verb!vertical_specialization!)
 \item \verb!leontief_output!; takes a decompr object and applies the Leontief decomposition using final output
 \item \verb!decomp!; a wrapper function which integrates the use of \verb!load_tables_vectors! with the various decompositions, using a argument \textit{method} to specify the desired decomposition (default \verb!leontief!) 
\end{itemize}

@@

The first step of the analytic process is to load the input object and create a decompr class object, 
which contains the data structures for the decompositions.
This step is not needed when using the \verb!decomp! wrapper function but more on this later.

@@

The application of the Leontief insight to ICIOs as opposed to national Input-Output tables for our Leontief decomposition is straightforward.
\textit{V} refers now to a vector of direct value added contributions of all industries across the different countries. 
Its dimension is correspondingly $1 x GN$, where $G$ is according to the number of countries.
\textit{A} is now of dimension $GN x GN$ and gives the industry flows including cross border relationships.
Since we are interested in the value added origins of exports we multiply these two matrices with a $GN x GN$ matrix whose diagonal we fill with each industry's exports, $E$, 
such that the basic equation behind the source decomposition is given by $V(I-A)^{-1}E$. 
\footnote{When using the leontief\_output function, the value added multiplier is instead multiplied with each industry's output.}
In a simple example with two countries (\textit{k} and \textit{l}) and industries (\textit{i} and \textit{j}) we can zoom in to see the matrices' content:

@@

\subsection{Implentation}
As described, in \cref{sec:data}, the first step of our analytic process is to construct a decompr object using the \verb!load_tables_vectors! function. After this, we can use the \verb!leontief! function to apply to Leontief decomposition.
<<leontief >>=
lt <- leontief( decompr_object )
@

In addition, a wrapper function called \verb!decomp! is provided which integrates both elements of the workflow into a single function.
We recomended that the atomic functions be used for large data sets, however, for small data sets this is an easy way to derive the results immediatly.
The \verb!decomp! function requires a method to be specified (see \verb!help("decomp")! for details), is none is provided, the function will default to \verb!leontief!.

@@

The first eight components relate to domestic value added of the exporting country contained in the sectoral exports of the exporting industry to the direct importer. 
For instance, the first non-zero element in \cref{tab:wwz} refers to $DVA\_FIN$, or domestic value added in final good exports. 
It shows that there are 5.47 units of domestic value added in the exports of final goods from Argentina's Agriculture industry to Turkey. 
In the same row the third term,  \textit{DVA\_INTrexI1}, is slightly more complicated. 
As mentioned above it gives the amount of domestic Value added in intermediate exports re-exported to third countries as intermediate goods to produce domestic final goods. 
In our example this means that there are 1.14 units of domestic value added in the intermediate exports of Argentina's Agriculture industry to Turkey, 
that are re-exported by Turkey as intermediates to a third country which produces final goods with it. 
Terms six to eight concern domestic value added that eventually returns home. 
\textit{RDV\_FIN2} reveals for example that there are 0.35 units of domestic value added in the intermediate exports of Argentina's Agriculture industry to Turkey, 
that Turkey re-exports as intermediates to Argentina for the latter's final good production.

The following four terms apply to foreign value added in exports and separate on the one hand between the origin of the foreign value added ($MVA$ vs $OVA$) and on the other hand between the type of export (intermediate vs final good). 
$MVA$ describes hereby foreign value added sourced by the exporting country from the direct importer. 
From the perspective of the latter, these terms are thus part of the $RDV$ (value added returning home) share. 
$OVA$ in contrast sums over the foreign value added sourced from all other countries. 
Going back to the example, this means that there 0.21 units of Turkish value added in the final goods exports of Argentina's Agriculture industry to Turkey.

Terms 13 to 16 collect the double counting of gross trade statistics that occurs when goods cross borders multiple times. 
$DDC$ captures double counting due to domestic value added, 
which is further classified according to the type of the ultimate export (final vs intermediate good). 
$MDC$ and $ODC$ on the other hand capture double counting due to foreign value added from either the direct importer or other countries. 
For the Argentina-Turkey case this implies, for instance, that there are 0.18 units of value added in the intermediate exports of Turkey to Argentina which are re-exported by Argentina's Agriculture industry to Turkey as intermediates and then again re-exported. 
Since they would be part of $MVA$ twice, they are now counted once as double-counted term.

@@

One interesting application of this decomposition for trade and development uses changes over time in $FVA\_FIN$ and $FVA\_INT$. 
When low-wage developing countries enter GVCs, they tend to specialize mainly in assembly but try to gradually move up within the value chain.    
To illustrate this, we can reuse the example of the iPhone. 
Most of the value added in the device stems from US design and Japanese technology but it is ultimately assmbled in China.
This means for China that when it enters this GVC, its $FVA\_FIN$ starts to increase since it imports a lot of foreign value added, assembles it, and exports a final good: the Iphone.
However, assembly itself does not contain a lot of value added so that the benefit of China initially is small. 
When its technology improves due to the interaction with Japan and the USA, it might be able to produce actual parts of the phone,
which contain more value added.
Eventually it might even be able to outsource assembly to a cheaper country.
This would imply that it seizes a larger share of the value added, something commonly referred to as upgrading or moving up within the value chain.
In terms of the WWZ decomposition, we would then observe that China's $FVA\_FIN$ first goes up and then starts to decline with
a simultaneous increase in $FVA\_INT$.

@@

We have designed the package using a modular structure, with an additional user interface function for increased ease of use.
The modular structure enables users to break the computationally intensive analysis process down into several steps.
Furthermore, the modular structure enables users and other developers to build on top of basic data structures which are creaed by the \verb!load_tables_vectors! function when implementing other decompositions or analyses.
However, a wrapper function (\verb!decomp!) is also provided, which combines the use of the atomic functions into one.
Lastly, in addition to the decompr function, we provide a Rcmdr plugin \parencite{fox2005getting, fox2007extending}, which allows the user to perform the analytical process using a GUI (Grapical User Interface).
All of this should allow users of the package to adapt the package to their specific needs as the GVC research progresses.

>>>>

\section{Introduction}
Global Value Chains (GVCs) refer to the quickly expanding internationalization of production networks.
Most goods we use nowadays consist of parts that are sourced from different corners of the planet 
and are assembled across different continents. A popular example of this development is the iPhone,
which uses inputs from at least five countries (USA, China, Germany, Taiwan, South Korea) and is assembled in two (USA and China).
This has made GVCs a central topic in research on trade and development policy.
Both policy makers and academia increasingly value the growth opportunities GVCs offer to global trade and, 
especially, to developing countries.
However, analysing this phenomenon empirically requires complex matrix manipulations, 
since the relevant data is only available in the form of gross flows.
The decompr package enables researchers with little background in matrix algebra and linear programming to easily derive standard GVC indicators for statistical analysis.

@@

\begin{itemize}
 \item \verb!load_tables_vectors!; transforms the input objects to an object used for the decompositions (class: decompr)
 \item \verb!leontief!; takes a decompr object and applies the Leontief decomposition
 \item \verb!wwz!; takes a decompr object and applies the Wang-Wei-Zhu decomposition
 \item \verb!vertical_specialisation!; takes a decompr object and returns a measure of vertical specialisation (also works with \verb!vertical_specialization!)
 \item \verb!leontief_output!; takes a decompr object and applies the Leontief decomposition using final output
 \item \verb!decomp!; a wrapper function which integrates the use of \verb!load_tables_vectors! with the various decompositions, using an argument \textit{method} to specify the desired decomposition (default \verb!leontief!) 
\end{itemize}

@@

The first step of the analytical process is to load the input object and create a decompr class object, 
which contains the data structures for the decompositions.
This step is not needed when using the \verb!decomp! wrapper function but more on this later.

@@

The application of the Leontief insight to ICIOs as opposed to national Input-Output tables for our Leontief decomposition is straightforward.
\textit{V} refers now to a vector of direct value added contributions of all industries across the different countries. 
Its dimension is correspondingly $1 x GN$, where $G$ is the number of countries.
\textit{A} is now of dimension $GN x GN$ and gives the industry flows including cross border relationships.
Since we are interested in the value added origins of exports we multiply these two matrices with a $GN x GN$ matrix whose diagonal we fill with each industry's exports, $E$, 
such that the basic equation behind the source decomposition is given by $V(I-A)^{-1}E$. 
\footnote{When using the leontief\_output function, the value added multiplier is instead multiplied with each industry's output.}
In a simple example with two countries (\textit{k} and \textit{l}) and industries (\textit{i} and \textit{j}) we can zoom in to see the matrices' content:

@@

\subsection{Implentation}
As described, in \cref{sec:data}, the first step of our analytical process is to construct a decompr object using the \verb!load_tables_vectors! function. After this, we can use the \verb!leontief! function to apply to Leontief decomposition.
<<leontief >>=
lt <- leontief( decompr_object )
@

In addition, a wrapper function called \verb!decomp! is provided which integrates both elements of the workflow into a single function.
We recomended that the atomic functions be used for large data sets, however, for small data sets this is an easy way to derive the results immediatly.
The \verb!decomp! function requires a method to be specified (see \verb!help("decomp")! for details), if none is provided, the function will default to \verb!leontief!.

@@

The first eight components relate to domestic value added of the exporting country contained in the sectoral exports of the exporting industry to the direct importer. 
For instance, the first non-zero element in \cref{tab:wwz} refers to $DVA\_FIN$, or domestic value added in final good exports. 
It shows that there are 5.47 units of domestic value added in the exports of final goods from Argentina's Agriculture industry to Turkey. 
In the same row the third term,  \textit{DVA\_INTrexI1}, is slightly more complicated. 
As mentioned above, it gives the amount of domestic Value added in intermediate exports re-exported to third countries as intermediate goods to produce domestic final goods. 
In our example this means that there are 1.14 units of domestic value added in the intermediate exports of Argentina's Agriculture industry to Turkey, 
that are re-exported by Turkey as intermediates to a third country which produces final goods with it. 
Terms six to eight concern domestic value added that eventually returns home. 
\textit{RDV\_FIN2} reveals for example that there are 0.35 units of domestic value added in the intermediate exports of Argentina's Agriculture industry to Turkey, 
that Turkey re-exports as intermediates to Argentina for the latter's final good production.

The following four terms apply to foreign value added in exports and separate on the one hand between the origin of the foreign value added ($MVA$ vs $OVA$) and on the other hand between the type of export (intermediate vs final good). 
$MVA$ describes hereby foreign value added sourced by the exporting country from the direct importer. 
From the perspective of the latter, these terms are thus part of the $RDV$ (value added returning home) share. 
$OVA$ in contrast sums over the foreign value added sourced from all other countries. 
Going back to the example, this means that there are 0.21 units of Turkish value added in the final goods exports of Argentina's Agriculture industry to Turkey.

Terms 13 to 16 collect the double counting of gross trade statistics that occurs when goods cross borders multiple times. 
$DDC$ captures double counting due to domestic value added, 
which is further classified according to the type of the ultimate export (final vs intermediate good). 
$MDC$ and $ODC$, on the other hand, capture double counting due to foreign value added from either the direct importer or other countries. 
For the Argentina-Turkey case this implies, for instance, that there are 0.18 units of value added in the intermediate exports of Turkey to Argentina which are re-exported by Argentina's Agriculture industry to Turkey as intermediates and then again re-exported. 
Since they would be part of $MVA$ twice, they are now counted once as double-counted term.

@@

One interesting application of this decomposition for trade and development uses changes over time in $FVA\_FIN$ and $FVA\_INT$. 
When low-wage developing countries enter GVCs, they tend to specialize mainly in assembly but try to gradually move up within the value chain.    
To illustrate this, we can reuse the example of the iPhone. 
Most of the value added in the device stems from US design and Japanese technology but it is ultimately assembled in China.
This means for China that when it enters this GVC, its $FVA\_FIN$ starts to increase since it imports a lot of foreign value added, assembles it, and exports a final good: the iPhone.
However, assembly itself does not contain a lot of value added so that the benefit of China initially is small. 
When its technology improves due to the interaction with Japan and the USA, it might be able to produce actual parts of the phone,
which contain more value added.
Eventually it might even be able to outsource assembly to a cheaper country.
This would imply that it seizes a larger share of the value added, something commonly referred to as upgrading or moving up within the value chain.
In terms of the WWZ decomposition, we would then observe that China's $FVA\_FIN$ first goes up and then starts to decline with
a simultaneous increase in $FVA\_INT$.

@@

We have designed the package using a modular structure, with an additional user interface function for increased ease of use.
The modular structure enables users to break the computationally intensive analysis process down into several steps.
Furthermore, the modular structure enables users and other developers to build on top of basic data structures which are created by the \verb!load_tables_vectors! function when implementing other decompositions or analyses.
However, a wrapper function (\verb!decomp!) is also provided, which combines the use of the atomic functions into one.
Lastly, in addition to the decompr function, we provide a Rcmdr plugin \parencite{fox2005getting, fox2007extending}, which allows the user to perform the analytical process using a GUI (Grapical User Interface).
All of this should allow users of the package to adapt the package to their specific needs as the GVC research progresses.

### 2015-02-10 draft.tex
stelina's comments

\subsection{Implentation}
As described, in \cref{sec:data}, the first step of our analytic process is to construct a decompr object using the \verb!load_tables_vectors! function. After this, we can use the \verb!leontief! function to apply to Leontief decomposition.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{lt} \hlkwb{<-} \hlkwd{leontief}\hlstd{( decompr_object )}
\end{alltt}
\end{kframe}
\end{knitrout}

In addition, a wrapper function called \verb!decomp! is provided which integrates both elements of the workflow into a single function.
We recomended that the atomic functions be used for large data sets, however, for small data sets this is an easy way to derive the results immediatly.
The \verb!decomp! function requires a method to be specified (see \verb!help("decomp")! for details), is none is provided, the function will default to \verb!leontief!.

>>>>

\subsection{Implentation}
As described, in \cref{sec:data}, the first step of our analytical process is to construct a decompr object using the \verb!load_tables_vectors! function. After this, we can use the \verb!leontief! function to apply to Leontief decomposition.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{lt} \hlkwb{<-} \hlkwd{leontief}\hlstd{( decompr_object )}
\end{alltt}
\end{kframe}
\end{knitrout}

In addition, a wrapper function called \verb!decomp! is provided which integrates both elements of the workflow into a single function.
We recomended that the atomic functions be used for large data sets, however, for small data sets this is an easy way to derive the results immediatly.
The \verb!decomp! function requires a method to be specified (see \verb!help("decomp")! for details), if none is provided, the function will default to \verb!leontief!.
